Full Title of Article

### 2022-06-14 acml/acml22_paper.tex
continued part of the structure

% givasile commands
\newcommand{\xc}{\mathbf{x}_c}
\newcommand{\Xc}{\mathbf{X}_c}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Ysloc}{Y_{s}^{\text{local}}}
\newcommand{\ysloc}{y_{s}^{\text{local}}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\xii}{\mathbf{x}^{(i)}}
\newcommand{\yi}{y^{(i)}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\1}{\mathbbm{1}}
\newcommand{\D}{\mathcal{D}}

@@

@@

\section{Probabilistic ALE plots}

\subsection{The ALE method}\label{sec:ALE}

In this section, we introduce the reader to the feature effect method
ALE. Given that \(f: \R^S \rightarrow \R \) is known, we can measure
the \textit{local} effect of the \(s\)-th feature at a specific point
\( \x = (\xc, x_s)\) of the input space \(\mathcal{X}\), as
\(f_s(\x) = \sfrac{\partial f(\x)}{\partial x_s}\). ALE models the
\textit{local} feature effect at \(x_s\) as an expectation over the
distribution of the unkown (latent) features
\(\E_{p(\Xc;x_s)}[f_s(x_s, \xc)]\). Afterwards, ALE measures the
\textit{global} effect at \(x_s\) as an accumulation of the expected
\textit{local} effects:
%
\begin{equation}
  \label{eq:ale-definition}
  f_{\mathtt{ALE}}(x_s) =
  \int_{x_{s, \text{min}}}^{x_s} \E_{p(\Xc;x_s=z)}[f_s(x_s, \xc)] \partial z
\end{equation}

In real cases, it is infeasible to compute
eq.~\eqref{eq:ale-definition} analytically. Therefore, we reside on
estimating the effect from the training set. Let's denote the
available dataset as \(\D = \{\xii, \yi \}_{i=1}^N\), where \( \xii \)
is the \(i\)-th feature vector of the training data and \(\yi\) is the
\(i\)-th label. Apley et. al proposed spliting the axis into \( K \)
equal-sized bins, find the set of points that lie in each bin, i.e.
\( \mathcal{S}_k = \{ \xii : x_s^{(i)} \in [z_{k-1}, z_k) \} \) and,
finally, find the local effect at each bin as the mean value of the
population, i.e.
\(\hat{\mu}_{s,k} = \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in
  \mathcal{S}_k} [f_s(\xii)]\). The global effect at \(x_s\) is then
estimated through:
%
\begin{equation}
  \label{eq:ale-approximation}
  \hat{f}_{\mathtt{ALE}}^s(x)
  = \Delta x \sum_{k=1}^{k_{x_s}} \hat{\mu}_{s,k}
  = \Delta x \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} [f_s(\xii)]
\end{equation}
%
where \(\Delta x\) is the bin size and \(k_{x_s}\) the bin-index of the value \(x_s\).
%
%
\subsection{ALE plots with uncertainty quantification}

ALE directly defines the local effect as an expectation over the
unknown features and the global effect as integral over the
expectations. Therefore, both local and global effects are modeled as
simple variables. In constrast, for integrating ALE plots in a
probabilistic framework, we model the local effect as a random
variable, dependent on the random variable \(\Xc\) representing the
latent features:

\begin{equation}
    \Ysloc \sim p(\ysloc ; x_s)
    = \int_{\xc} \delta(\ysloc-f_s(\x)) p(\xc;x_s) \partial \xc
    \label{eq:local-effect-rv}
\end{equation}
%
Following the core idea of ALE, we define the global feature effect as
the random variable \(Y_s\) which is the integration of the local
effects:

\begin{equation}
  Y_s \sim p(y_s ; x_s) = \int_{x_{s,min}}^{x_s} p(\ysloc;x_s=z) \partial z
  \label{eq:global-effect-rv}
\end{equation}
%
Through eqs.~(\ref{eq:local-effect-rv}), (\ref{eq:global-effect-rv})
we formulate the ALE in a fully probabilistic manner. We notice, that
the definition of ALE as given by~\cite{Apley2020} is simply the
expected value of \(Y_s\), i.e.
\(f_{\mathtt{ALE}}(x_s) = \E[Y_s ; x_s]\). For simpler notation we
denote \(\E[Y_s ; x_s]\) as \(\mu_s (x_s)\).

For measuring the uncertainty, it is also important to capture the
variance of the global effect. Notating the variance of the feature
effect \(\var[Y_s ; x_s]\) as \(\sigma_s^2\), we can compute it
through:

\begin{equation}
  \sigma_s^2 (x_s)
  = \int_{x_{s,min}}^{x_s} \int_{\xc} p(\xc|x_s) (f_s(\x) - \mu_s(x_s))^2 \partial \xc (\partial z)^2
\end{equation}

As before, it is infeasible to compute the the variance
\(\sigma_s^2(x_s)\) in closed-form. Therefore, we reside on estimating
them from the available examples of the training set. The estimation
of the variance can also done through the available points:

\begin{equation}
  \label{eq:effect-var-approx}
  \hat{\sigma}_s^2(x_s)
  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \hat{\sigma}_{s,k}^2
  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} (f_s(\xii) - \hat{\mu}_{s,k})^2 
\end{equation}

@@

@@

\subsection{Variable-size bins as an optimization problem}

@@

@@

\subsection{Subsection Title}

@@

%\bibliographystyle{plain}
\bibliography{acml22}

@@

\section{Derivations}\label{apd:first}

@@

>>>>

% givasile commands
\newcommand{\xc}{\mathbf{x}_c}
\newcommand{\Xc}{\mathbf{X}_c}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Ysloc}{Y_{s}^{\text{local}}}
\newcommand{\ysloc}{y_{s}^{\text{local}}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\xii}{\mathbf{x}^{(i)}}
\newcommand{\xsi}{x_s^{(i)}}
\newcommand{\xci}{\mathbf{x}_c^{(i)}}
\newcommand{\yi}{y^{(i)}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\1}{\mathbbm{1}}
\newcommand{\D}{\mathcal{D}}

@@

The decision about the bin size has major influence in the feature
effect plot. Therefore, it is very important (a) to inform the user to
what extend they should trust the ALE plot and, consequently, decide
the bin size with the optimal effect. In the end, we will see that it i

\newpage
~\newpage

@@

\newpage

\section{Probabilistic formulation of ALE plots}

\subsection{ALE background}\label{sec:ALE}

This section introduces the reader to the ALE formulation of feature
effect. Given that \(f: \R^S \rightarrow \R \) is known, we can
measure the effect of the \(s\)-th feature at a specific point
\( \x = (\xc, x_s)\) of the input space \(\mathcal{X}\) as
\(f_s(\x) = \sfrac{\partial f(\x)}{\partial x_s}\). ALE models the
\textit{local} effect at \(x_s\) as the expected effect over the
distribution of the unkown (latent) features \(\Xc\), i.e.
\(\E_{p(\xc;x_s)}[f_s(x_s, \xc)]\). Afterwards, the \textit{global}
effect at \(x_s\) is the accumulation of the \textit{local} effects:
%
\begin{equation}
  \label{eq:ale-definition}
  f_{\mathtt{ALE}}(x_s) =
  \int_{x_{s, \text{min}}}^{x_s} \E_{p(\xc;x_s=z)}[f_s(x_s, \xc)] \partial z
\end{equation}

In real cases, it is infeasible to compute
eq.~\eqref{eq:ale-definition} analytically. Therefore, we reside on
estimating the effect from the training set. Let's denote the
available dataset as \(\D = \{\xii, \yi \}_{i=1}^N\), where \( \xii \)
is the \(i\)-th feature vector of the training data and \(\yi\) is the
\(i\)-th label.~\cite{Apley2020} proposed spliting the axis into
\( K \) equal-sized bins, find the set of points that lie in each bin,
i.e.  \( \mathcal{S}_k = \{ \xii : x_s^{(i)} \in [z_{k-1}, z_k) \} \)
and, finally, find the local effect at each bin as the mean value of
the population \(\hat{\mu}_{s,k}\). The global effect is then
estimated through the following formula:
%
\begin{equation}
  \label{eq:ale-approximation}
  \hat{f}_{\mathtt{ALE}}(x_s)
  = \Delta x \sum_{k=1}^{k_{x_s}} \hat{\mu}_{s,k}
  = \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xsi \in \mathcal{S}_k} [f(z_k, \xci) - f(z_{k-1}, \xci)]
\end{equation}
%
where \(\Delta x\) is the bin size and \(k_{x_s}\) the index of the
bin that includes \(x_s\).
%

\subsection{Quantification of the uncertainty of the estimation}

We propose an alternative estimation of the feature effect inside each
bin. The alternative estimation is better because it (a) separates the
decision about the bin limits from the local effect of each point, (b)
secures from out-of-distribution sampling and (c) provides better
variance estimation inside the bin. The alternative estimation is:

\begin{equation}
  \label{eq:1}
  \tilde{f}_{\texttt{ALE}}(x_s) =
  \Delta x \sum_{k=1}^{k_{x_s}} \tilde{\mu}_{s,k} =
  \Delta x \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xsi \in \mathcal{S}_k} f_s(\xii)
\end{equation}
%
Our goal is to inform the user to what extent they should trust the
ALE plot. For this purpose, we introduce two metrics; the variance and
the standard error of the estimation. The variance of the estimation
is:

\begin{equation}
  \label{eq:effect-var-approx}
  \tilde{\sigma}_s^2(x_s)
  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \tilde{\sigma}_{s,k}^2
  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} (f_s(\xii) - \tilde{\mu}_{s,k})^2 
\end{equation}
%
The standard error of the estimation is:

\begin{equation}
  \label{eq:effect-std-error}
  \text{SE}(x_s) = \Delta x \sum_{k=1}^{k_{x_s}} (\frac{\hat{\sigma}_{s,k}^2}{|\mathcal{S}_k|})^\frac{1}{2}
\end{equation}

The two metrics, (\ref{eq:effect-var-approx}) and
(\ref{eq:effect-std-error}), should be trusted only in case they
respect some contstraints. Under these constraints, the user can be
confident that the correct feature effect is inside two or three
standard errors errors around the mean estimation.

Our computations are based on the hypothesis that inside
all bins \textbf{the gradient wrt. to the feature of interest doesn't
  depend on the value feature of interest}. Unfortunately, we cannot
when this is the case. We just know that as the bins grow larger, it
is more possible to violated this hypothesis. In this case both the
ALE effect and the standard error are wrong.

The standard error should be trusted when it is estimated by a large
of data points. For example, in plots (c) and (d), there are bins with
less than 10 points. Therefore, in these cases, we cannot trust the
plot or the standard error.

@@

Based on the observations of the previous chapter, we want to create a
single metric for choosing the best feature effect plot. Our goal is
to minimize the variance, given that we have points inside eac bin

We propose the minimization of the accumulated standard deviation (or
accumulated variance). For ALE with K bins, let's notate as:

\begin{itemize}
\item \(dx^K\) the length of each bin
\item \(p_i^K \) the number of the training points inside the \(i-th\) bin
\item \(\sigma_i^K \) the std of the local effects of the training points inside the \(i-th\) bin
\end{itemize}
%
We will minimize:

\begin{align} \label{eq:min-criterion}
  K_{min} &= \text{argmin}_{K} \quad  [dx^K  \sum_i^K \sigma_i^K * (1 - d_i^K)]\\
  & \text{s.t. } p_i^K \geq \texttt{min\_points\_per\_bin } \forall i
\end{align}
%
where \(d_i^K = 0.2 * \frac{p_i^K}{N} \in [0,0.1] \) works as a `discount', favoring the
creation of bigger bins in cases of similar standard deviation.

@@

\newpage
~\newpage

@@

\begin{enumerate}
\item One example to show that (a) variance and (b) standard error are
  good estimates to what extend to trust the ALE plot.
\item One example to show the importance of variable size bins
\end{enumerate}

\newpage
~\newpage

\newpage

@@

\newpage

@@

%\bibliographystyle{plain}
\bibliography{acml22}
\newpage

@@

\section{Symbols}\label{apd:symbols}

Spaces and points:

\begin{itemize}
\item \(\mathcal{X} \subseteq \R^S \): the input space
\item \(\x = (x_s, \xc) \in \mathcal{X}\)
\item \(x_s \in \R \)
\item \(\xc \in \R^{S-1}\)
\item
\end{itemize}
%
Functions:
\begin{itemize}
\item \(f: \R^S \rightarrow \R\): black-box function
\item \(f_s(\x) = \sfrac{\partial f(\x)}{\partial x_s}: \R^S \rightarrow \R\): effect at point \(\x\)
\item 
\end{itemize}
%
ALE:
\begin{itemize}
\item \(\sfrac{f(z_{k_x}) - f(z_{k_x-1})}{\Delta x}\): effect of \(x_s\) at \(\x\)
\item \(\E_{p(\xc;x_s)}[f_s(x_s, \xc)]\): local effect of \(x_s\) at \(x_s\)
\item \(f_{\texttt{ALE}}(x_s): \R \rightarrow \R\): global effect of \(x_s\) at \(x_s\)
\item \(\hat{f}_{\texttt{ALE}}(x_s): \R \rightarrow \R\): \textit{estimator} of global effect of \(x_s\) at \(x_s\)
\end{itemize}

\section{Derivations}\label{apd:derivations}

@@

% \subsection{Probabilistic formulation of ALE plots}

% In this section, we will introduce a probabilistic formulation for ALE
% plots. As discussed before, ALE directly defines the local effect as
% the expectated value over the unknown features,
% i.e. \(\E_{p(\xc;x_s)}[f_s(x_s, \xc)]\). Therefore, the local and the
% global effect, which is the integration of the local effects, are
% modeled as deterministic variables.

% For adjusting ALE plots into a probabilistic framework, we model the
% local effect with a random variable conditioned on feature \(x_s\),
% i.e., \(\Ysloc \sim p(\ysloc ; x_s)\):

% \begin{equation}
%     p(\ysloc ; x_s)
%     = \int_{\xc} \delta(\ysloc-f_s(\x)) p(\xc;x_s) \partial \xc
%     \label{eq:local-effect-rv}
% \end{equation}
% %
% Following the core idea of ALE, we define the global feature effect as
% the accumulation of the local effects. Therefore, the feature effect
% of the \(s\)-th feature is \(Y_s \sim p(y_s ; x_s) \):
% \begin{equation}
%   p(y_s ; x_s) = \int_{x_{s,min}}^{x_s} p(\ysloc;x_s=z) \partial z
%   \label{eq:global-effect-rv}
% \end{equation}
% %
% In eqs.~(\ref{eq:local-effect-rv}), (\ref{eq:global-effect-rv}) we
% reformulate ALE plots in a fully probabilistic manner. We are
% particularly interested into two summary statistics of \(Y_s\), the
% expected value \(\mu_s (x_s) = \E[Y_s ; x_s]\) and the variance
% \(\sigma_s^2(x_s) = \var[Y_s ; x_s]\). In the appendix, we show that
% the expected value coincides with the definition of ALE plots in
% eq.~(\ref{eq:ale-definition}), i.e.
% \(\mu_s (x_s) = f_{\texttt{ALE}(x)s}\). The variance can be computed
% as:

% \begin{equation}
%   \sigma_s^2 (x_s)
%   = \int_{x_{s,min}}^{x_s} \E_{p(\xc|x_s = z)}[(f_s(\x) - \mu_s(z))^2](\partial z)^2
% \end{equation}

% As before, it is infeasible to compute the the variance
% \(\sigma_s^2(x_s)\) in analytically. Therefore, we reside on
% estimating them from the available examples of the training set:

% \begin{equation}
%   \label{eq:effect-var-approx}
%   \hat{\sigma}_s^2(x_s)
%   = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \hat{\sigma}_{s,k}^2
%   = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} (f_s(\xii) - \hat{\mu}_{s,k})^2 
% \end{equation}

% .. add a description for standard error

% \begin{equation}
%   \label{eq:effect-std-error}
%   \text{SE}(x_s) = \Delta x \sum_{k=1}^{k_{x_s}} (\frac{\hat{\sigma}_{s,k}^2}{|\mathcal{S}_k|})^\frac{1}{2}
% \end{equation}
